STA302 Fall 2024 Final Project Part 1: Research Proposal and Data Introduction

### 2024-12-06 final_paper/final_report.Rmd
refined model

# Contributions
- **Gadiel David Flores:** Description of contributions.
- **Wendy Huang:** Description of contributions.

@@

\newpage 
# Methods

library(knitr)
library(here)
library(kableExtra)
library(broom)

@@

This section shows the results of the multiple linear regression (MLR) analysis. The response variable for this analysis was New Housing Price Index (NHPI), and 13 predictors were selected, representing factors such as absorption rates, GDP, CPI, and various housing construction metrics. These predictors were chosen based on their theoretical relevance to housing price trends and their potential statistical significance.

Although removing influential observations (rows 1 and 4) improved model fit and precision, we decided against excluding them to avoid overfitting and to maintain the integrity of the dataset.

@@

Key predictors such as CPI, Unabsorbed Homes, and Under Construction Homes emerged as significant contributors to the model, with consistently low p-values and strong relationships with the response variable. However, issues of multicollinearity remain evident, as demonstrated by high Variance Inflation Factor (VIF) values for predictors like CPI, GDP and Detached Absorption. Multicollinearity inflates standard errors, potentially reducing the reliability of coefficient estimates, which warrants further refinement, such as variable selection or dimensionality reduction. For this reason, the GDP predictor was removed due to its high VIF and lack of significance, along with Completed Construction Semi and Starting Semi Construction.

Transformations applied to the predictors, such as logarithmic adjustments to Detached Absorption Quarterly Average, Starting Detached Construction, and Detached Unabsorbed Quarterly Average, addressed non-linear relationships observed in the scatterplots, improving the linearity assumption of regression. The removal of predictors like GDP Quarterly Average, Starting Semi Construction, and Semi Unabsorbed Quarterly Average was motivated by their non-significant p-values and high multicollinearity, streamlining the model while retaining its predictive power. Additionally, the application of a Box-Cox transformation to the response variable mitigated issues of skewness and heteroscedasticity, ensuring constant variance across the fitted values.

Residual diagnostic plots indicate that the model reasonably satisfies the key assumptions of linear regression. Residuals are well-centered around zero, with no clear patterns in residuals versus fitted values, suggesting that the linearity and homoscedasticity assumptions are mostly satisfied. However, minor deviations at the tails in the Q-Q plot and some patterns in predictor-specific residual plots hint at potential areas for improvement, such as exploring alternative transformations or addressing influential points.

@@

# Generate a LaTeX table for PDF
kable(model_summary, 
      format = "latex", 
      caption = "Model Coefficients and Summary",
      col.names = c("Term", "Estimate", "Std. Error", "t value", "p value")) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  column_spec(5, bold = TRUE, color = "black")

>>>>

# Contributions
- **Gadiel David Flores:** Methods, flowchart and Results
- **Wendy Huang:** Description of contributions.

@@

\newpage 
# Methods
\setlength{\parindent}{2em} 
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(knitr)
library(here)
library(kableExtra)
library(broom) 
library(dplyr)
library(gridExtra)
library(grid)
library(png)

@@

This section presents the results of the multiple linear regression (MLR) analysis. The response variable for this analysis was the New Housing Price Index (NHPI), and eight predictors were selected to represent factors such as absorption, interest rates, GDP, CPI, and various housing construction metrics. These predictors were chosen based on their theoretical relevance to housing price trends and their potential statistical significance. The focus of this model is explanatory, as the aim is to understand the relationships between variables.

We begin by identifying influential observations. Although removing influential observations (rows 1 and 4) improved model fit and precision, we decided against excluding them to avoid overfitting and to maintain the integrity of the dataset.

@@

Key predictors such as CPI, Unabsorbed Homes, and Completed Construction Homes emerged as significant contributors to the model, with consistently low p-values and strong relationships with the response variable. However, issues of multicollinearity remain evident, as demonstrated by high Variance Inflation Factor (VIF) values for predictors like CPI, GDP and Detached Absorption. Multicollinearity inflates standard errors, potentially reducing the reliability of coefficient estimates, which warrants further refinement, such as variable selection or dimensionality reduction. For this reason, the GDP predictor was removed due to its high VIF and lack of significance, along with starting Construction Homes, and Under Construction Homes as they were not of high significance.

Transformations applied to the predictors, such as logarithmic adjustments to Absorption, Completed Construction, and Unabsorbed homes, addressed non-linear relationships observed in the scatterplots, improving the linearity assumption of regression. The removal of predictors like GDP, and Starting Construction was motivated by their non-significant p-values and high multicollinearity, streamlining the model while retaining its predictive power. Additionally, the application of a Box-Cox transformation to the response variable mitigated issues of skewness and heteroscedasticity, ensuring constant variance across the fitted values.

img1 <- rasterGrob(png::readPNG(here::here("final_paper", "completed_contruction.png")), interpolate = TRUE)
img2 <- rasterGrob(png::readPNG(here::here("final_paper", "cpi.png")), interpolate = TRUE)
img3 <- rasterGrob(png::readPNG(here::here("final_paper", "aborption.png")), interpolate = TRUE)
img4 <- rasterGrob(png::readPNG(here::here("final_paper", "resid_v_fitted.png")), interpolate = TRUE)

grid.arrange(img1, img2, img3, img4, ncol = 2)
```

Residual diagnostic plots indicate that the model reasonably satisfies the key assumptions of linear regression. Residuals are well-centered around zero, with no clear patterns in residuals versus fitted values, suggesting that the linearity and homoscedasticity assumptions are mostly satisfied.

After applying all the necessary transformations to address non-linearity and improve model fit, the updated regression model demonstrated strong performance metrics. The \( R^2 \) value of 0.9742 indicates that 97.42% of the variance in the response variable is explained by the predictors, with an adjusted \( R^2 \) of 0.9712 accounting for the number of predictors used. The model's Akaike Information Criterion (AIC) of 237.3722 and Bayesian Information Criterion (BIC) of 247.5055 suggest a well-optimized model with minimal overfitting. Additionally, the Root Mean Squared Error (RMSE) of 4.048 confirms a low average prediction error, indicating that the transformations significantly improved the model's predictive accuracy.

@@

model_summary <- model_summary %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

kable(model_summary, 
      format = "latex", 
      caption = "Model Coefficients and Summary",
      col.names = c("Term", "Estimate", "Std. Error", "t value", "p value")) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  column_spec(5, bold = TRUE, color = "black")

### 2024-12-06 final_paper/paper.qmd
refined model

# response variable and predictors
response <- "Quarterly_Average"
predictors <- c(
  "Detached_Absorption_Quarterly_Avg",
  "Semi_Absorption_Quarterly_Avg",
  "Detached_Unabsorbed_Quarterly_Avg",
  "Semi_Unabsorbed_Quarterly_Avg",
  "GDP_Quarterly_Avg",
  "CPI_Quarterly_Avg",
  "Starting_Detached_Construction",
  "Starting_Semi_Construction",
  "Under_Construction_Detached",
  "Under_Construction_Semi",
  "Completed_Construction_Semi",
  "Completed_Construction_Detached",
  "rates"
)

@@

# Transform predictors to address non-linearity
train_data$Detached_Absorption_Quarterly_Avg <- log(train_data$Detached_Absorption_Quarterly_Avg + 1)
train_data$Starting_Detached_Construction <- log(train_data$Starting_Detached_Construction + 1)
train_data$Under_Construction_Detached <- log(train_data$Under_Construction_Detached + 1)
train_data$Completed_Construction_Semi <- NULL 
train_data$Detached_Unabsorbed_Quarterly_Avg <- log(train_data$Detached_Unabsorbed_Quarterly_Avg + 1)
train_data$Semi_Unabsorbed_Quarterly_Avg <- NULL 
train_data$Starting_Semi_Construction <- NULL

# Box-Cox transformation for response variable to address skewness and heteroscedasticity
bc <- boxcox(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg +
                                Starting_Detached_Construction +
                                Detached_Unabsorbed_Quarterly_Avg +
                                Under_Construction_Detached +
                                CPI_Quarterly_Avg,
             data = train_data, plotit = FALSE)
lambda <- bc$x[which.max(bc$y)]

@@

# Fit the linear regression model
model <- lm(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg +
                                   Starting_Detached_Construction +
                                   Detached_Unabsorbed_Quarterly_Avg +
                                   Under_Construction_Detached +
                                   CPI_Quarterly_Avg,
            data = train_data)

@@

predictors <- c(
  "Detached_Absorption_Quarterly_Avg",
  "Detached_Unabsorbed_Quarterly_Avg",
  "GDP_Quarterly_Avg",
  "CPI_Quarterly_Avg",
  "Starting_Detached_Construction",
  "Under_Construction_Detached",
  "Completed_Construction_Detached",
  "rates"
)

@@

## 6. Model Validation

>>>>

# response variable and predictors
response <- "Quarterly_Average"
predictors <- c(
  "Detached_Absorption_Quarterly_Avg",
  "Detached_Unabsorbed_Quarterly_Avg",
  "GDP_Quarterly_Avg",
  "CPI_Quarterly_Avg",
  "Starting_Detached_Construction",
  "Under_Construction_Detached",
  "Completed_Construction_Detached",
  "rates"
)

@@

# Transform predictors for non-linearity
train_data$Detached_Absorption_Quarterly_Avg <- log(train_data$Detached_Absorption_Quarterly_Avg + 1)
train_data$CPI_Quarterly_Avg <- log(train_data$CPI_Quarterly_Avg + 1)
train_data$Completed_Construction_Detached <- log(train_data$Completed_Construction_Detached + 1)

# Box-Cox transformation for response variable to address skewness and heteroscedasticity
bc <- boxcox(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg +
               Detached_Unabsorbed_Quarterly_Avg +
               Completed_Construction_Detached +
                                CPI_Quarterly_Avg,
             data = train_data, plotit = FALSE)
lambda <- bc$x[which.max(bc$y)]

@@

# Fit the linear regression model
model <- lm(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg +
                                    +
                                   Detached_Unabsorbed_Quarterly_Avg +
                                   Completed_Construction_Detached +
                                   CPI_Quarterly_Avg,
            data = train_data)

@@

predictors <- c(
  "Detached_Absorption_Quarterly_Avg",
  "Detached_Unabsorbed_Quarterly_Avg",
  "CPI_Quarterly_Avg",
  "Completed_Construction_Detached"
)

@@

## Model Validation
